### 2025-06-24 sections/2-methods.tex
reorganizing and moving to supplement

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
    \REQUIRE $f: X \times X \mapsto \mathbb{R}^+ \cup \{0\}$, a distance function
    \REQUIRE $C$, a cluster
    \REQUIRE $criteria$, user-specified continuation criteria

    \STATE $seeds \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
    \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
    \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
    \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

    \IF{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
        \STATE Partition($L$, $criteria$)
    \ENDIF

    \IF{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
        \STATE Partition($R$, $criteria$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

When building a tree for a real dataset, the depth would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of $\mathcal{T}$ would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
We do, however, explore empirical results based on forcing a balanced tree in Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations}.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($\mathcal{R}$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $\mathcal{R}$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find
        \STATE{$H \Leftarrow [\ ]$, a max-heap by $\delta$ of size $k$}
        \STATE $r \Leftarrow radius$ of $\mathcal{R}$
        \STATE $r \Leftarrow$ $\frac{r}{|\mathcal{R}|}$
        \STATE $Q \Leftarrow$ tree-search($\mathcal{R}$, $q$, $r$)
        \WHILE{$\sum_{C \in Q} |C| < k$}
            \IF{$Q = \emptyset$}
                \STATE $r \Leftarrow 2 \cdot r$
            \ELSIF{$\sum_{C \in Q} |C| >= k$}
                \STATE $\mu \Leftarrow \frac{1}{|Q|} \cdot \sum_{C \in Q} \big( LFD(C)^{-1} \big)$
                \STATE $r \Leftarrow r \cdot \min \bigg( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu} \bigg)$
            \ENDIF
            \STATE $Q \Leftarrow$ tree-search($\mathcal{R}$, $q$, $r$)
        \ENDWHILE
        \STATE $H \Leftarrow \text{leaf-search}(Q, q, r)$
        \STATE \textbf{return} $H$ as a list
    \end{algorithmic}
\end{algorithm}

@@

The work in~\cite{yu2015entropy} showed that $\sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d$, where $\gamma$ is a constant.
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus, $\sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d$.
It remains to determine an estimate for $\rho_k$.
We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at the center of a cluster in $Q$.
By Equation~\ref{eq:methods:lfd-original}, $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$.
We can rearrange this equation to get $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get:
\begin{equation*}
    k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d
\end{equation*}

@@

    \begin{algorithm}
        \caption{Breadth-First Sieve($\mathcal{R}$, $q$, $k$)} % give the algorithm a caption
        \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
        \begin{algorithmic} % enter the algorithmic environment
            \REQUIRE $\mathcal{R}$, the root cluster
            \REQUIRE $q$, a query
            \REQUIRE $k$, the number of neighbors to find
            \STATE $c \Leftarrow$ \textit{center} of $\mathcal{R}$
            \STATE $Q \Leftarrow$ \{ ($\mathcal{R}$, $\delta^{+}_{\mathcal{R}}$, $|\mathcal{R}| - 1$), ($c$, $\delta_{\mathcal{R}}$, 1) \}
            \WHILE{$\sum_{(\_, \_, m) \in Q} m \neq k$}
                \STATE $\tau \Leftarrow$ QuickSelect($Q$, $k$)
                \STATE $Q^{'} \Leftarrow \emptyset$
                \FOR{$(C, \_, \_) \in Q$}
                    \IF{$\delta^{-}_{C} \leq \tau$}
                        \IF{$C$ is a point}
                            \STATE $Q^{'} \Leftarrow Q^{'} \cup \{ (C, \delta_{C}, 1) \}$
                        \ELSIF{$C$ is a leaf}
                            \STATE $Q^{'} \Leftarrow Q^{'} \cup \{ (p, \delta_{p}, 1)$ for $p \in C \}$
                        \ELSE
                            \STATE $[L, R] \Leftarrow$ children of $C$
                            \STATE $l, r \Leftarrow$ centers of $L, R$
                            \STATE $Q^{'} \Leftarrow Q^{'} \cup \{ (L, \delta^{+}_{L}, |L| - 1), (l, \delta_{L}, 1) \}$
                            \STATE $Q^{'} \Leftarrow Q^{'} \cup \{ (R, \delta^{+}_{R}, |R| - 1), (l, \delta_{R}, 1) \}$
                        \ENDIF
                    \ENDIF
                \ENDFOR
                \STATE $Q \Leftarrow Q^{'}$
            \ENDWHILE
            \STATE QuickSelect($Q$, $k$)
            \STATE \textbf{return} The first $k$ points in $Q$
        \end{algorithmic}
    \end{algorithm}

    \begin{algorithm}
        \caption{Depth-First Sieve($\mathcal{R}$, $q$, $k$)}
        \label{alg:methods:depth-first-sieve}
        \begin{algorithmic}
            \REQUIRE $\mathcal{R}$, the root cluster
            \REQUIRE $q$, a query
            \REQUIRE $k$, the number of neighbors to find
            \STATE{$Q \Leftarrow [\mathcal{R}]$, a min-heap by $\delta^{-}$}
            \STATE{$H \Leftarrow [\ ]$, a max-heap by $\delta$ of size $k$}
            \WHILE{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
                \WHILE{$Q.peek$ is not a leaf}
                    \STATE{$C \Leftarrow Q.pop$, the closest cluster}
                    \STATE{$[L, R] \Leftarrow$ children of $C$}
                    \STATE{$Q.push(L)$}
                    \STATE{$Q.push(R)$}
                \ENDWHILE
                \STATE{$leaf \Leftarrow Q.pop$}
                \FOR{$p \in leaf$}
                    \STATE{$H.push(p)$}
                \ENDFOR
            \ENDWHILE
            \STATE \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}

\vskip 0.1in

>>>>

% \begin{algorithm} % enter the algorithm environment
%     \caption{Partition($C$, $criteria$)} % give the algorithm a caption
%     \label{alg:methods:partition} % and a label for \ref{} commands later in the document
%     \begin{algorithmic} % enter the algorithmic environment
%     \REQUIRE $f: X \times X \mapsto \mathbb{R}^+ \cup \{0\}$, a distance function
%     \REQUIRE $C$, a cluster
%     \REQUIRE $criteria$, user-specified continuation criteria

%     \STATE $seeds \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
%     \STATE $c \Leftarrow$ geometric median of $seeds$
%     \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
%     \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
%     \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
%     \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

%     \IF{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
%         \STATE Partition($L$, $criteria$)
%     \ENDIF

%     \IF{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
%         \STATE Partition($R$, $criteria$)
%     \ENDIF
% \end{algorithmic}
% \end{algorithm}

@@

% When building a tree for a real dataset, the depth would be larger than $\log n$, which would seem to increase the cost of tree-building.
% However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of $\mathcal{T}$ would also have fewer points, and so the cost of building each subsequent level would be lower.
% Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
% We do, however, explore empirical results based on forcing a balanced tree in Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations}.
% In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

@@

The work in~\cite{yu2015entropy} showed that 
\begin{equation} \sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d \end{equation}, where $\gamma$ is a constant.
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k.$
Thus, $\sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d$.
It remains to determine an estimate for $\rho_k$.
We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at the center of a cluster in $Q$.
By Equation~\ref{eq:methods:lfd-original}, $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$.
We can rearrange this equation to get $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get:
\begin{equation*}
    k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d
\end{equation*}

@@

% \vskip 0.1in

### 2025-06-24 sections/3-datasets.tex
formatting fixes
reorganizing and moving to supplement

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
Table~\ref{tab:datasets:summary} summarizes these datasets.
All benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.164-1-MANJARO.
The Rust compiler was Rust 1.83.0, and the Python interpreter version was 3.9.18.

\begin{table}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \begin{center}
        \begin{sc}
            \begin{tabular}{|l|l|l|l|}
                \hline
                \textbf{Dataset} & \textbf{Distance Function}  &\textbf{Cardinality}  & \textbf{Dimensionality}  \\
                \hline
                Fashion-Mnist    & Euclidean                   & 60,000             & 784                    \\
                \hline
                Glove-25         & Cosine                      & 1,183,514          & 25                     \\
                \hline
                Sift             & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                Random           & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                SILVA            & Levenshtein                 & 2,224,640          & 3,712         \\
                \hline
                RadioML          & Dynamic Time Warping        & 97,920             & 1,024                  \\
                \hline
            \end{tabular}
        \end{sc}
    \end{center}
    \vskip -0.1in
\end{table}

@@

>>>>

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
Table~\ref{tab:datasets:summary} summarizes the cardinality and dimensionality of each these datasets.
All benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.164-1-MANJARO.
The Rust compiler was Rust 1.83.0, and the Python interpreter version was 3.9.18.

\begin{table}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \begin{center}
        \begin{sc}
            \begin{tabular}{|l|l|l|l|}
                \hline
                \textbf{Dataset} & \textbf{Dist. Function}  &\textbf{Card}  & \textbf{Dim}  \\
                \hline
                Fashion-Mnist    & Euclidean                   & 60,000             & 784                    \\
                \hline
                Glove-25         & Cosine                      & 1,183,514          & 25                     \\
                \hline
                Sift             & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                Random           & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                SILVA            & Levenshtein                 & 2,224,640          & 3,712         \\
                \hline
                RadioML          & Dynamic Time Warping        & 97,920             & 1,024                  \\
                \hline
            \end{tabular}
        \end{sc}
    \end{center}
    \vskip -0.1in
\end{table}

@@

\subsection{Local Fractal Dimension}
\label{sec:dayasets:lfd-of-datasets}

Since the CAKES algorithms were designed to scale in time with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and Radio-ML.
In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.

For the Fashion-MNIST dataset we observe in Figure~\ref{fig:results:fashion-mnist-lfd} that until approximately depth 5, Fashion-MNIST's LFD is low (i.e., less than 4).
It then starts increasing, reaching a peak of about 6 near depth 20, before decreasing to 1 at the maximum depth.

Relative to Fashion-MNIST, Glove-25 has low LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}.
In particular, Glove-25's LFD is less than 3 for all depths.

Figure~\ref{fig:results:sift-lfd} shows the LFD by depth for Sift, which has higher LFD relative to Fashion-MNIST and Glove-25.
It increases sharply to a peak of 9 around a depth of 10, after which it decreases smoothly until reaching the deepest leaves in the tree.

We contrast the low LFD of Sift with that of the Random dataset which has the same cardinality and dimension. As shown in Figure~\ref{fig:results:random-lfd}, the character of this dataset is very different from the others.
The LFD starts at 20 at depth 0 and all percentile lines decrease linearly with depth until reaching the leaves of the tree.
The spread in LFD starts very small for the first few clusters and increases as depth increases.

The Silva dataset, shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
In particular, LFD is less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

For the Radio-ML dataset, the LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
Within each of the tree portions, this dataset has a character similar to that of the Random dataset.
This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth.
This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.
